### 2021-11-22 manuscript/hiv-gwas.tex
Update on Overleaf.

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{bm} % for bold math characters
\usepackage{booktabs} % for prettier hlines in table
\usepackage{subcaption} % for subfigures
\usepackage[affil-it]{authblk} % for footnote-style author affiliation
\usepackage{setspace} % for 1.5-spaced lines not including footnotes, floats
\usepackage{tikz} % for graphical model drawing
\usepackage{float} % to enforce figure positioning in subsection they're declared in
\usepackage{amsmath}
\usepackage[singlelinecheck=false]{caption} % to left-justify single-line table captions
\usepackage{array}
\newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}} % for centered, width-specified columns 
\usepackage{makecell} % for multi-row table entries
\usepackage{tabularx} % for tables that don't spill past margin

\usepackage{url} % otherwise bibtex hates my references file
\usepackage[round,sort]{natbib}

@@

\startsetups
    \setcounter{figure}{0}
    \renewcommand{\thefigure}{S\arabic{figure}}
    \setcounter{table}{0}
    \renewcommand{\thetable}{S\arabic{table}}
\stopsetups

@@

\begin{flushright}
Submission type: Article\\
Section: Methods
\end{flushright}

@@

\noindent
$^{1}$Department of Biosystems Science and Engineering, ETH Zürich, Basel, Switzerland\\
$^{2}$Swiss Institute of Bioinformatics, Lausanne, Switzerland\\ 
$^{3}$Precision Medicine Unit, Lausanne University Hospital and University of Lausanne, Lausanne, Switzerland\\ 
$^{4}$Institute of Medical Virology, University of Zurich, Zurich, Switzerland\\ 
$^{5}$Division of Infectious Diseases and Hospital Epidemiology, University Hospital Zurich, University of Zurich, Zurich, Switzerland\\
$^{6}$Division of Infectious Diseases, Laboratory of Virology, Geneva University Hospital, Geneva, Switzerland\\
$^{7}$Division of Immunology and Allergy, University Hospital Lausanne, Lausanne, Switzerland\\
$^{8}$Department of Biomedicine, University of Basel, Basel, Switzerland\\
$^{9}$Department of Infectious Diseases, Bern University Hospital and University of Bern, Bern, Switzerland\\
$^{10}$Division of Clinical Virology, University Hospital Basel, Basel, Switzerland\\
$^{11}$Division of Infectious Diseases and Hospital Epidemiology, University Hospital Basel, Basel, Switzerland\\
$^{12}$Division of Infectious Diseases, University Hospital Lausanne, Lausanne, Switzerland\\
$^{13}$Division of Infectious Diseases, Cantonal Hospital St. Gallen, St. Gallen, Switzerland\\
$^{14}$Division of Infectious Diseases, Regional Hospital Lugano, Lugano, Switzerland\\
$^{15}$Global Health Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne, Lausanne, Switzerland\\
$\dagger$Co-last authors\\
$*$Corresponding author: tanja.stadler@bsse.ethz.ch

\newpage
\begin{doublespace}

\begin{abstract}
	Infectious diseases are a unique challenge for genome-wide association studies (GWAS) because pathogen, host, and environmental factors can all effect disease traits. Previous GWAS have successfully identified several human genetic variants associated with HIV-1 set point viral load (spVL), among other important infectious disease traits. However, these GWAS do not account for potentially confounding or extraneous pathogen effects that are heritable from donor to recipient in transmission chains.  We propose a novel method to consider the full genome of each patient's infecting pathogen strain, remove strain-specific effects on a trait based on the pathogen phylogeny, and thus better estimate the effect of human genetic variants on infectious disease traits. In simulations, we show our method can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations. When we apply our method to HIV-1 cohort data from the Swiss HIV Cohort Study, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatibility complex (MHC) gene regions compared to standard GWAS. Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 spVL, standard GWAS are robust to pathogen effects. Our framework may improve GWAS for other diseases where pathogen effects are stronger and more phylogenetically correlated amongst individuals in a cohort.
\end{abstract}

@@

% Why is spVL and important trait
For HIV, GWAS have used a trait called set point viral load (spVL) to identify human variants associated with severity of disease course. spVL is generally defined to be the average concentration of viral particles in host plasma during the asymptomatic phase of infection in the absence of treatment \citep{Mellors1996}. In untreated individuals, spVL is predictive of duration of asymptomatic infection \citep{Mellors1996} and infectiousness \citep{Quinn2000}. If viral load can be reduced to undetectable levels, an individual is effectively uninfectious and the risk of disease progression is massively reduced \citep{HHSARTGuidelines2019}. Notably, spVL varies by orders of magnitude between individuals \citep{Mellors1996}. Thus, spVL measurements point to a wide range in natural HIV control amongst individuals.

% How do we measure the extent to which genetic factors determine spVL
To understand HIV pathogenicity, it is important to understand to what extent spVL is determined by host genetic factors \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}. Heritability is a key measure of how genetically-determined a trait is. Here we distinguish between two different heritability measures that are used in different contexts in infectious disease study. Broad-sense heritability $H^2$ measures the fraction of total trait variance that is heritable, i.e. due to inherited differences. In the infectious disease case, broad-sense heritability from pathogen factors, which are inherited by recipients from their infection partners, is typically measured. On the other hand, narrow-sense heritability $h^2$ measures the fraction of total trait variance due specifically to additive genetic effects, i.e. the sum of independent effects from all genetic variants. GWAS for infectious disease typically measure narrow-sense heritability of a trait based on human genetic variants.

@@

% How heritable is spVL (viral side)
In addition to these human genetic factors, it is well-recognized that viral genetic factors effect spVL. As mentioned, heritability from the viral side is typically measured using broad-sense heritability. Estimates differ depending on the methods employed and the cohort studied (see \citet{Mitov2018} for a discussion of this uncertainty). Estimates using phylogenetic methods on large UK and Swiss cohorts by \citep{Mitov2018} and \citep{Bertels2018} measured the broad-sense heritability of spVL from the virus to be 21\% - 29\%. However, variation in the MHC is known to exert strong selective pressure on the virus \citep{Kloverpris2016, Nguyen2021SystematicLink}. If the virus can change its ``puzzle piece'' shape to escape MHC-presentation, infected cells can go undetected. This means that MHC variants affect spVL largely via selection on the virus \citep{Bartha2017}. In summary, human genetic factors play a role in determining spVL, but these effects may be mediated by viral genetic variation.

% The gap
Most of the GWAS for human genetic determinants of spVL conducted so far \citep{Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012, McLaren2015} do not explicitly consider any viral effect on spVL. In these GWAS, viral genetic effects are lumped in with residual variance due to other, non-genetic factors. This has several potential negative consequences, namely that (i) viral effects may be confounding or extraneous variables that bias estimates of host genetic effects, (ii) variability due to viral effects would make it more challenging to identify human variants of small effect, and (iii) spVL values from a cohort are not truly independent samples, given that patients closer in the transmission chain have more similar strains and therefore more similar spVL values.

% Overview of relevant methods and why we don't use them
Issue (iii) is closely related to a well-known problem in standard GWAS, namely that shared (human) ancestry, especially between close relatives, can give rise to spurious genetic correlations with a trait. Corrections for these correlations are well-developed and widely accepted  \citep{Astle2009, Price2006}. More recently, \citep{Power2016} emphasized the need to do similar corrections for shared pathogen ancestry in microbial GWAS. Two state-of-the-art methods exist for this \citep{Earle2016, collins_phylogenetic_nodate}. 
% \citep{Earle2016} simultaneously test for pathogen lineage and locus-level associations with a trait using a linear mixed model framework. \citep{collins_phylogenetic_nodate} test for locus-level associations that are significantly stronger than expected under a null model considering shared pathogen ancestry. 
However, these approaches are only suitable to quantify effects from $pathogen$ genetic variants on a trait. In contrast, we want to estimate effects of $human$ genetic variants on a trait, taking into account pathogen effects. \citep{Naret2018} developed a relevant method for this task. The authors suggest adding principle components derived from the pathogen phylogeny as covariates to the linear regression models for association testing. This should correct for trait correlations due to shared pathogen ancestry. However, the top principle components capture only some of the information from the full pathogen phylogeny. Furthermore, we would like to simultaneously address issues (i) and (ii).

% Introduction to our method
In this work,  we draw from the field of phylogenetic comparative methods to develop a new GWAS framework that estimates and removes trait variability due to the pathogen using information from the full pathogen phylogeny. Our approach should help identify human genetic variants that affect disease traits and more accurately estimate their effects. Furthermore, we think this approach represents the first GWAS framework to identify human genetic variants associated with a disease trait that accounts for non-independence of samples due to the underlying pathogen phylogeny.

In the following we describe a statistical model for the spVL trait, derive a maximum likelihood estimate for the viral part of spVL under this model, and describe a new infectious disease GWAS framework using this information. In simulations, we demonstrate that this framework can improve GWAS power to detect host genetic variants effecting a disease trait. Finally, we apply our framework to human and viral genome data from the Swiss HIV Cohort Study (SHCS) and show that associations with spVL are robust to a correction for viral effects. Although we developed our framework in the context of HIV-1 spVL, the same approach could readily be applied to other heritable infectious disease traits.

@@

Variation in spVL comes from several sources: direct host genetic effects, pathogen effects, interaction effects between the host and the pathogen, and other environmental effects. Of these, only pathogen effects are heritable from one transmission partner to another \citep{Leventhal2016}. To characterize these effects, we use a phylogenetic mixed model (PMM) \citep{Housworth2004}. PMMs assume continuous traits like spVL are the sum of independent heritable and non-heritable parts. In our case, pathogen effects comprise the heritable part and all other effects comprise the non-heritable part. Then, the heritable part is modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny, as in Figure \ref{fig:spVLModel}A. The non-heritable part is modeled as Gaussian noise added to sampled individuals at the tips of the phylogeny.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width = \linewidth]{figures/simulation_setup.pdf}
	\caption{A high-level schematic of our POUMM-based simulation framework. (A) shows how pathogen genetic effects on spVL evolve along the pathogen phylogeny according to an Ornstein Uhlenbeck process. (B) shows host genetic effects, which are the sum of independent effects from several causal variants. Each variant can be present in 0, 1, or 2 copies and we assume half have effect size $\delta$ on the trait value and half have effect size $-\delta$. (C) shows that environmental effects are assumed to be independent between sampled individuals and are drawn from a distribution centered at zero. Measured trait values are the sum of these three effects for each simulated individual.}
		\label{fig:spVLModel}
	\end{center}
\end{figure}

So far, PMMs with two types of random processes have been used to model spVL evolution. The Brownian Motion (BM) process assumes unbounded trait values, i.e. spVL can attain any value. The Ornstein-Uhlenbeck (OU) process assumes trait values fluctuate around an optimal value, i.e. extreme spVL values are unlikely. \citep{Mitov2018} and \citep{Bertels2018} previously showed the OU process has higher statistical support for spVL. This makes sense given that spVL is likely under stabilizing selection to maximize viral transmission potential \citep{Fraser2014}. Therefore, we assume the OU process. The full model is called the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) and is described in detail by \citet{Mitov2018}. Here, we review the main relevant points in the spVL context.

Under the POUMM, the spVL trait $z$ is the sum of viral effects $g_v$, host genetic effects $g_h$, and other environmental or interaction effects $\epsilon$. We can group the non-heritable effects $g_h$ and $\epsilon$ into a broader category of ``environmental'' effects $e$: 
\begin{equation}
    z = g_{v} + e
\end{equation}
$g_v$ is a viral trait that evolves along the phylogeny according to an OU process. The OU process is defined by a stochastic differential equation with two terms. The first term represents a deterministic pull towards an optimal trait value and the second term represents stochastic fluctuations modelled by Brownian motion \citep{Butler2004}:
\begin{align}
\begin{split}
	&dg_v(t) = \alpha[\theta - g_v(t)]dt + \sigma dW_t \\
	&g_v(0) = g_0 
	\label{eq:OUprocess}
\end{split}
\end{align}

Here the parameter $\alpha$ represents selection strength towards an evolutionarily optimal value represented by parameter $\theta$. The parameter $\sigma$ measures the intensity of stochastic fluctuations in the evolutionary process. Finally, $dW_t$ is the Wiener process underlying Brownian motion. The OU process is a Gaussian process, meaning that $g_v(t)$ is a Gaussian random variable. Assuming $g_v(t)$ starts at initial value $g_0$ at time $t = 0$ at the root of the phylogeny, we can write the expectation for $g_v(t)$ at time $t$: 
\begin{equation}
   E[g_v(t)] = g_0e^{-\alpha t} + (1 - e^{-\alpha t})\theta \label{eq:OUmean}
\end{equation}
and the variance in $g_v(t)$ if we were to repeat the random evolutionary process many times \citep{Butler2004}:
\begin{equation}
    Var[g_v(t)] = \frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t}) \label{eq:OUVar}
\end{equation}

$g_v$ evolves independently in descendent lineages after a divergence event in the phylogeny. The covariance between $g_v(t)$ in a lineage $i$ at time $t_i$ and another lineage $j$ at time $t_j$, $Cov\big(g_{v_i}(t_i), g_{v_j}(t_j)\big)$, increases with the amount of time between $t_0$ and the divergence of the two lineages, $t_{0(ij)}$, and decreases with the total amount of time the lineages evolve independently, $d_{ij}$ \citep{Butler2004}: 
\begin{equation}
	Cov\big(g_{v_i}(t_i), g_{v_j}(t_j)\big) = \frac{\sigma^2}{2\alpha}[e^{-\alpha d_{ij}}(1 - e^{-2\alpha t_{0(ij)}})]
	\label{eq:OUcov}
\end{equation}
Next, we remember that $e$ is the non-heritable, environmental part of spVL. $e$ is modeled as a Gaussian random variable that is time- and phylogeny-independent. The expectation of $e$ is 0, meaning environmental effects are equally likely to raise or lower spVL from the virus-determined level. The parameter $\sigma_e^2$ measures the between-host variance of the environmental effect.
\begin{align}
\begin{split}
	E(e) = 0 \\
	Var(e) = \sigma^2_e
\end{split}
\end{align}

Finally, broad-sense trait heritability can be calculated as the fraction of total trait variance that is heritable:
\begin{equation}
	H^2_{\bar{t}} = \frac{Var[g_v(t)]}{Var[g_v(t)] + Var(e)} = \frac{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t})}{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t}) + \sigma^2_e}
\end{equation}

@@

Considering that $\bm{G_v}$ and $\bm{E}$ are independent random vectors and that their realizations $\bm{g}_v$ and $\bm{e}$ must sum together to equal the observed spVL values $\bm{z}$, we can write the following proportionality for the joint probability density of $\bm{g}_v$ and $\bm{e}$:
\begin{equation}
	f\big(\bm{g}_v, \bm{e}\big) \propto \mathcal{N}\big(\bm{g}_v; \bm{\mu}_{G}, \boldsymbol{\Sigma}_G\big)
	\label{eq:pdfGprop}
\end{equation}

where the expected value of $\bm{g}_v$ and the covariance matrix $\boldsymbol{\Sigma}_G$ are defined as:
\begin{align}
	Exp(\bm{g}_v) = \bm{\mu}_{G} &=  \boldsymbol{\Sigma}_G\big(\boldsymbol{\Sigma}_{OU}^{-1}\bm{\mu}_{OU} + \boldsymbol{\Sigma}_E^{-1} \bm{z}\big) \label{eq:MuG}\\
	\boldsymbol{\Sigma}_G &= \big(\boldsymbol{\Sigma}_{OU}^{-1} + \boldsymbol{\Sigma}_E^{-1}\big)^{-1} \label{eq:SigmaG}
\end{align}

@@

	Equations \ref{eq:MuG} and \ref{eq:SigmaG} follow from eq. \ref{eq:pdfG} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \citep{Petersen2012}.
\end{proof}

@@

Given the estimator we just derived for $\bm{g}_v$, we can now estimate $\bm{e}$, the spVL value $without$ pathogen effects: 
\begin{equation}
	\hat{\bm{e}} = \bm{z} - Exp(\bm{g}_v)
	\label{eq:EHat}
\end{equation}

@@

To test the theoretical best-case performance of our method, we simulated data under the POUMM and applied our framework to the simulated data. Figure \ref{fig:spVLModel} shows a high-level schematic of our simulation framework and Table \ref{tab:sim-params} gives the value or expression for each parameter. In a nutshell, we simulated independent, additive host genetic effects, independent environmental effects, and heritable pathogen genetic effects under different scenarios of trait heritability and selection strength. To maintain the same heritability while varying selection strength, we counter-balanced by varying the intensity of stochastic evolutionary fluctuations accordingly. We fixed other variables to plausible values based on the spVL literature.

@@

Human genotypes, viral load measurements, and HIV-1 $pol$ gene sequences from HIV-1 positive individuals were all collected in the context of other studies by the Swiss HIV Cohort Study (SHCS, www.shcs.ch, \citet{Schoeni-Affolter2010, Scherrer2021CohortSHCS}). All participants were HIV-1–infected individuals at least 16 years old and written informed consent was obtained from all cohort participants. The anonymized data was made available for this study after a study proposal was approved by the SHCS.

>>>>

\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{lineno} % for line numbers
\usepackage{graphicx}
\usepackage{bm} % for bold math characters
\usepackage{booktabs} % for prettier hlines in table
\usepackage{subcaption} % for subfigures
\usepackage[affil-it]{authblk} % for footnote-style author affiliation
\usepackage{setspace} % for 1.5-spaced lines not including footnotes, floats
\usepackage{tikz} % for graphical model drawing
\usepackage{float} % to enforce figure positioning in subsection they're declared in
\usepackage{amsmath}
\usepackage[singlelinecheck=false]{caption} % to left-justify single-line table captions
\usepackage{array}
\newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}} % for centered, width-specified columns 
\usepackage{makecell} % for multi-row table entries
\usepackage{tabularx} % for tables that don't spill past margin

\usepackage{url} % otherwise bibtex hates my references file
\usepackage[round,sort]{natbib}
\linespread{1.3}

@@

@@

% \begin{flushright} % for MBE submission
% Submission type: Article\\
% Section: Methods
% \end{flushright}

@@

\noindent
$^{1}$Department of Biosystems Science and Engineering, ETH Zurich, Basel, Switzerland\\
$^{2}$Swiss Institute of Bioinformatics, Lausanne, Switzerland\\ 
$^{3}$Precision Medicine Unit, Lausanne University Hospital and University of Lausanne, Lausanne, Switzerland\\ 
$^{4}$Institute of Medical Virology, University of Zurich, Zurich, Switzerland\\ 
$^{5}$Division of Infectious Diseases and Hospital Epidemiology, University Hospital Zurich, University of Zurich, Zurich, Switzerland\\
$^{6}$Division of Infectious Diseases, Laboratory of Virology, Geneva University Hospital, Geneva, Switzerland\\
$^{7}$Division of Immunology and Allergy, University Hospital Lausanne, Lausanne, Switzerland\\
$^{8}$Department of Biomedicine, University of Basel, Basel, Switzerland\\
$^{9}$Department of Infectious Diseases, Bern University Hospital and University of Bern, Bern, Switzerland\\
$^{10}$Division of Clinical Virology, University Hospital Basel, Basel, Switzerland\\
$^{11}$Division of Infectious Diseases and Hospital Epidemiology, University Hospital Basel, Basel, Switzerland\\
$^{12}$Division of Infectious Diseases, University Hospital Lausanne, Lausanne, Switzerland\\
$^{13}$Division of Infectious Diseases, Cantonal Hospital St. Gallen, St. Gallen, Switzerland\\
$^{14}$Division of Infectious Diseases, Regional Hospital Lugano, Lugano, Switzerland\\
$^{15}$Global Health Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne, Lausanne, Switzerland\\
$\dagger$Co-last authors\\
$*$Corresponding author: tanja.stadler@bsse.ethz.ch

\newpage
\begin{linenumbers}
% \begin{doublespace}

\begin{abstract}
	Infectious diseases are a unique challenge for genome-wide association studies (GWAS) because pathogen, host, and environmental factors can all affect disease traits. Previous GWAS have successfully identified several human genetic variants associated with HIV-1 set point viral load (spVL), among other important infectious disease traits. However, these GWAS do not account for potentially confounding or extraneous pathogen effects that are heritable from donor to recipient in transmission chains.  We propose a new method to consider the full genome of each patient's infecting pathogen strain, remove strain-specific effects on a trait based on the pathogen phylogeny, and thus better estimate the effect of human genetic variants on infectious disease traits. In simulations, we show our method can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations. When we apply our method to HIV-1 subtype B data from the Swiss HIV Cohort Study, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatibility complex (MHC) gene regions compared to standard GWAS. Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 subtype B spVL, standard GWAS are robust to pathogen effects. Our framework may improve GWAS for other diseases if pathogen effects are even more phylogenetically correlated amongst individuals in a cohort.
\end{abstract}

@@

% Why is spVL and important trait
For HIV, GWAS have used a trait called set point viral load (spVL) to identify human variants associated with the severity of disease course. spVL is generally defined to be the average concentration of viral RNA copies in host plasma during the asymptomatic phase of infection in the absence of treatment (see e.g. \cite{Alizon2010}). In untreated individuals, spVL is predictive of the duration of asymptomatic infection \citep{Mellors1996} and infectiousness \citep{Quinn2000}. If viral load can be reduced to undetectable levels, an individual is effectively uninfectious and the risk of disease progression is massively reduced \citep{HHSARTGuidelines2019}. Notably, spVL varies by orders of magnitude between individuals \citep{Mellors1996}. Thus, spVL measurements point to a wide range in natural HIV control amongst individuals.

% How do we measure the extent to which genetic factors determine spVL
To understand HIV pathogenicity, it is important to understand to what extent spVL is determined by host genetic factors \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}. Heritability is a key measure of how genetically-determined a trait is. Here we distinguish between two different heritability measures that are used in different contexts in the study of infectious diseases. Broad-sense heritability $H^2$ measures the fraction of total trait variance that is heritable, i.e. due to inherited differences. In the infectious disease case, broad-sense heritability from pathogen factors, which are inherited by recipients from their infection partners, is typically measured. On the other hand, narrow-sense heritability $h^2$ measures the fraction of total trait variance due specifically to additive genetic effects, i.e. the sum of independent effects from all genetic variants. GWAS for infectious disease traits typically measure the narrow-sense heritability of a trait based on human genetic variants.

@@

% How heritable is spVL (viral side)
In addition to these human genetic factors, it is well-recognized that viral genetic factors affect spVL. As mentioned, heritability from the viral side is typically measured using broad-sense heritability. Estimates differ depending on the methods employed and the cohort studied (see \citet{Mitov2018} for a discussion of this uncertainty). Estimates using phylogenetic methods on large UK and Swiss cohorts by \citet{Mitov2018} and \citet{Bertels2018} measured the broad-sense heritability of spVL from the virus to be 21\% - 29\%. However, variation in the MHC is known to exert strong selective pressure on the virus \citep{Kloverpris2016, Nguyen2021SystematicLink}. If the virus can change its ``puzzle piece'' shape to escape MHC-presentation, infected cells can go undetected. This means that MHC variants affect spVL largely via selection on the virus \citep{Bartha2017}. In summary, human genetic factors play a role in determining spVL, but these effects may be due to interaction with specific viral genetic variants.

% The gap
Most of the GWAS for human genetic determinants of spVL conducted so far \citep{Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012, McLaren2015} do not explicitly consider any viral effect on spVL. In these GWAS, viral genetic effects are lumped in with residual variance due to other, non-genetic factors. This has several potential negative consequences. (i) Viral effects may be confounding or extraneous variables that bias estimates of host genetic effects. (ii) Variability due to viral effects would make it more challenging to identify human variants of small effect. Finally, (iii) spVL values from a cohort are not truly independent samples, given that patients closer in the transmission chain have more similar strains and therefore more similar spVL values.

% Overview of relevant methods and why we don't use them
Issue (iii) is closely related to a well-known problem in standard GWAS. Shared (human) ancestry, especially between close relatives, can also give rise to spurious genetic correlations with a trait. Corrections for these correlations are well-developed and widely accepted  \citep{Astle2009, Price2006}. More recently, \citet{Power2017} emphasized the need to do similar corrections for shared pathogen ancestry in microbial GWAS. Two state-of-the-art methods exist for this \citep{Earle2016, collins_phylogenetic_nodate}. 
% \citep{Earle2016} simultaneously test for pathogen lineage and locus-level associations with a trait using a linear mixed model framework. \citep{collins_phylogenetic_nodate} test for locus-level associations that are significantly stronger than expected under a null model considering shared pathogen ancestry. 
However, these approaches are only suitable to quantify effects from $pathogen$ genetic variants on a trait. In contrast, we want to estimate effects of $human$ genetic variants on a trait, while accounting for pathogen effects. \citet{Naret2018} developed a relevant method for this in the context of a genome-to-genome GWAS framework. The authors suggest adding principle components derived from the pathogen phylogeny as covariates to the linear regression models for association testing. This should correct for trait correlations due to shared pathogen ancestry. However, the top principle components capture only some of the information from the full pathogen phylogeny. Furthermore, we would like to simultaneously address issues (i) and (ii).

% Introduction to our method
In this work,  we draw from the field of phylogenetic comparative methods to develop a new GWAS framework that estimates and removes trait variability due to the pathogen using information from the full pathogen phylogeny. Our approach should help identify human genetic variants that affect disease traits and more accurately estimate their effects.

In the following we describe a statistical model for the spVL trait, derive a maximum likelihood estimate for the viral part of spVL under this model, and describe a new infectious disease GWAS framework using this information. In simulations, we show that this framework can improve GWAS power to detect host genetic variants that affect disease traits. Finally, we apply our framework to human and viral genome data from the Swiss HIV Cohort Study (SHCS) and show that associations with spVL are robust to a correction for viral effects. Although we developed our framework in the context of HIV-1 spVL, this approach can readily be applied to other heritable infectious disease traits.

@@

Variation in spVL comes from several sources: direct host genetic effects, pathogen effects, interaction effects between the host and the pathogen, and other environmental effects. Of these, only pathogen effects are heritable from one transmission partner to another \citep{Leventhal2016}. To characterize these effects, we use a phylogenetic mixed model (PMM) \citep{Housworth2004}. PMMs assume continuous traits like spVL are the sum of independent heritable and non-heritable parts. In our case, pathogen effects comprise the heritable part and all other effects comprise the non-heritable part. The heritable part is modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny, as in Figure \ref{fig:spVLModel}A. The non-heritable part is modeled as Gaussian noise added to sampled individuals at the tips of the phylogeny.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
	\caption{A high-level schematic of our POUMM-based simulation framework. (A) shows how pathogen genetic effects on spVL evolve along the pathogen phylogeny according to an Ornstein-Uhlenbeck process. (B) shows how host genetic effects are the sum of independent effects from several causal variants. Each variant can be present in 0, 1, or 2 copies. Half the variants have a positive effect of size $\delta$ and half have a negative effect of size $\delta$. (C) shows how environmental effects are independently drawn from a Gaussian distribution centered at 0. These three effects sum to the trait value for each simulated individual.}
		\label{fig:spVLModel}
	\end{center}
\end{figure}

So far, PMMs with two types of random processes have been used to model spVL evolution. The Brownian Motion (BM) process assumes unbounded trait values, i.e. spVL can attain any value. The Ornstein-Uhlenbeck (OU) process assumes trait values fluctuate around an optimal value, i.e. extreme spVL values are unlikely. \citet{Mitov2018} and \citet{Bertels2018} previously showed the OU process has higher statistical support for spVL. This makes sense given that spVL is likely under stabilizing selection to maximize viral transmission potential \citep{Fraser2014}. Therefore, we assume the OU process. The full model is called the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) and is described in detail by \citet{Mitov2018}. Here, we review the main points in the spVL context.

Under the POUMM, the spVL trait $z$ is the sum of viral effects $g_v$, host genetic effects $g_h$, and other environmental or interaction effects $\epsilon$. We can group the non-heritable effects $g_h$ and $\epsilon$ into a broader category of ``environmental'' effects $e$:

\begin{equation}
    z = g_{v} + e
\end{equation}
$g_v$ is a viral trait that evolves along the phylogeny according to an OU process. The OU process is defined by a stochastic differential equation with two terms. The first term represents a deterministic pull towards an optimal trait value and the second term represents stochastic fluctuations modelled by Brownian motion \citep{Butler2004}:

\begin{align}
\begin{split}
	&dg_v(t) = \alpha[\theta - g_v(t)]dt + \sigma dW_t \\
	&g_v(0) = g_0 
	\label{eq:OUprocess}
\end{split}
\end{align}

Here the parameter $\alpha$ represents selection strength towards an evolutionarily optimal value represented by parameter $\theta$. The parameter $\sigma$ measures the intensity of stochastic fluctuations in the evolutionary process. Finally, $dW_t$ is the Wiener process underlying Brownian motion. The OU process is a Gaussian process, meaning that $g_v(t)$ is a Gaussian random variable. Assuming $g_v(t)$ starts at initial value $g_0$ at time $t = 0$ at the root of the phylogeny, we can write the expectation for $g_v(t)$ at time $t$:

\begin{equation}
   E[g_v(t)] = g_0e^{-\alpha t} + (1 - e^{-\alpha t})\theta \label{eq:OUmean}
\end{equation}
and the variance in $g_v(t)$ if we were to repeat the random evolutionary process many times \citep{Butler2004}:
\begin{equation}
    Var[g_v(t)] = \frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t}) \label{eq:OUVar}
\end{equation}

$g_v$ evolves independently in descendent lineages after a divergence event in the phylogeny. The covariance between $g_v(t)$ in a lineage $i$ at time $t_i$ and another lineage $j$ at time $t_j$, $Cov\big(g_{v_i}(t_i), g_{v_j}(t_j)\big)$, increases with the amount of time between $t_0$ and the divergence of the two lineages, $t_{0(ij)}$, and decreases with the total amount of time the lineages evolve independently, $d_{ij}$ \citep{Butler2004}:

\begin{equation}
	Cov\big(g_{v_i}(t_i), g_{v_j}(t_j)\big) = \frac{\sigma^2}{2\alpha}[e^{-\alpha d_{ij}}(1 - e^{-2\alpha t_{0(ij)}})]
	\label{eq:OUcov}
\end{equation}
Next, we remember that $e$ is the non-heritable, environmental part of spVL. $e$ is modeled as a Gaussian random variable that is time- and phylogeny-independent. The expectation of $e$ is 0, meaning environmental effects are equally likely to raise or lower spVL from the virus-determined level. The parameter $\sigma_e^2$ measures the between-host variance of the environmental effect.

\begin{align}
\begin{split}
	E(e) = 0 \\
	Var(e) = \sigma^2_e
\end{split}
\end{align}

Finally, broad-sense trait heritability can be calculated as the fraction of total trait variance that is heritable:

\begin{equation}
	H^2_{\bar{t}} = \frac{Var[g_v(t)]}{Var[g_v(t)] + Var(e)} = \frac{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t})}{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t}) + \sigma^2_e}
\end{equation}

@@

Considering that $\bm{G_v}$ and $\bm{E}$ are independent random vectors and that their realizations $\bm{g}_v$ and $\bm{e}$ must sum together to equal the observed spVL values $\bm{z}$, we can write the following proportionality for the joint probability density of $\bm{g}_v$ and $\bm{e}$:

\begin{equation}
	f\big(\bm{g}_v, \bm{e}\big) \propto \mathcal{N}\big(\bm{g}_v; \bm{\mu}_{G}, \boldsymbol{\Sigma}_G\big)
	\label{eq:pdfGprop}
\end{equation}

where the expected value of $\bm{g}_v$ and the covariance matrix $\boldsymbol{\Sigma}_G$ are defined as:

\begin{align}
	Exp(\bm{g}_v) = \bm{\mu}_{G} &=  \boldsymbol{\Sigma}_G\big(\boldsymbol{\Sigma}_{OU}^{-1}\bm{\mu}_{OU} + \boldsymbol{\Sigma}_E^{-1} \bm{z}\big) \label{eq:MuG}\\
	\boldsymbol{\Sigma}_G &= \big(\boldsymbol{\Sigma}_{OU}^{-1} + \boldsymbol{\Sigma}_E^{-1}\big)^{-1} \label{eq:SigmaG}
\end{align}

@@

	Equations \ref{eq:MuG} and \ref{eq:SigmaG} follow from eq. \ref{eq:pdfG} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \citet{Petersen2012}.
\end{proof}

@@

Given the estimator we just derived for $\bm{g}_v$, we can now estimate $\bm{e}$, the spVL value $without$ pathogen effects:

\begin{equation}
	\hat{\bm{e}} = \bm{z} - Exp(\bm{g}_v)
	\label{eq:EHat}
\end{equation}

@@

To test the theoretical best-case performance of our method, we simulated data under the POUMM and applied our framework to the simulated data. Figure \ref{fig:spVLModel} shows a high-level schematic of our simulation framework and Table \ref{tab:sim-params} gives the value or expression for each parameter. For a description of the full simulation scheme, see Figure \ref{fig:sim-design}. In a nutshell, we simulated independent, additive host genetic effects, independent environmental effects, and heritable pathogen genetic effects under different scenarios of trait heritability and selection strength. To maintain the same heritability while varying selection strength, we counter-balanced by varying the intensity of stochastic evolutionary fluctuations accordingly. We fixed other variables to plausible values based on the spVL literature.

@@

Human genotypes, viral load measurements, and HIV-1 $pol$ gene sequences from HIV-1 positive individuals were all collected in the context of other studies by the Swiss HIV Cohort Study (SHCS) (www.shcs.ch, \citet{Schoeni-Affolter2010, Scherrer2021CohortSHCS}). All participants were HIV-1–infected individuals 16 years or older and written informed consent was obtained from all cohort participants. The anonymized data were made available for this study after the study proposal was approved by the SHCS.
